---
title: "workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SpFut.flexiSDM)
library(blockCV)
```

Hi! If you've found our package, you probably want to fit some species data into an integrated species distribution model. Throughout this vignette, we'll do just that. For demonstration purposes, let's assume that you want to model some data on the XX toad/frog/salamander.

### Setup

There are quite a few setup steps that must be done before you start using the package.

These are...


#### Setup model variables

```{r, eval=F}

# Set up model variables ----
mods <- read.csv("code/03-species-models/MVPv1.csv") %>% filter(number %in% nums.do)

tmp <- expand.grid(block.out = block, number = unique(mods$number))
mods <- full_join(mods, tmp, by = c("number"))

nums.do <- 8
block <- c("none")
local <- 1
a <- 1

# Get variables for model from MVPv1.csv
number <- mods$number[a]
sp.code <- mods$sp.code[a]
model <- mods$model[a]

sp.auto <- mods$sp.auto[a]
coarse.grid <- mods$coarse.grid[a]
year.start <- mods$year.start[a]
year.end <- mods$year.end[a]
buffer <- mods$buffer[a]
filter.region <- mods$filter.region[a]
spat.bal <- mods$spat.bal[a]
coordunc <- mods$coordunc[a]
coordunc_na.rm <- mods$coordunc_na.rm[a]
block.folds <- mods$block.folds[a]
block.rows <- mods$block.rows[a]
block.cols <- mods$block.cols[a]
block.out <- mods$block.out[a]

if (block.out == "none") {
  blockname <- "full"
} else {blockname <- block.out}

zero_mean <- 1

iter <- mods$iter[a]
thin <- mods$thin[a]
burnin <- floor(iter*0.75)
region.sub <- mods$region.sub[a]
lat.hi <- mods$lat.hi[a]
lat.lo <- mods$lat.lo[a]
lon.hi <- mods$lon.hi[a]
lon.lo <- mods$lon.lo[a]

project <- mods$project[a]
if (block.out != "none") {
  project <- 0
}

# Get covariates from MVPv1.csv
covs.PO <- unlist(str_split(mods$covs.PO[a], pattern = ", "))
covs.inat <- unlist(str_split(mods$covs.inat[a], pattern = ", "))
covs.lin <- unlist(str_split(mods$covs.lin[a], pattern = ", "))
covs.quad <- unlist(str_split(mods$covs.quad[a], pattern = ", "))
covs.int.factor <- unlist(str_split(mods$covs.int.factor[a], pattern = ", "))
reference <- mods$reference[a]
covs.int.cont <- unlist(str_split(mods$covs.int.cont[a], pattern = ", "))
check.covs <- mods$check.covs[a]

Bpriordist <- mods$Bpriordist[a]
Bpriorvar1 <- mods$Bpriorvar1[a]
Bpriorvar2 <- mods$Bpriorvar2[a]

process.intercept <- F
```

You are now ready to start processing your data.




### Data processing

```{r, eval=F}
common <- read.csv("DATA SWAMP/00-speciescodes.csv") %>%
          filter(DS.code == sp.code) %>%
          pull(DS.common)


# make output file
out.dir <- paste0("outputs/03-species-models/MVPv1/", number, "_", sp.code, "_", model, "/")
if (dir.exists(out.dir) == F) {
  dir.create(out.dir)
}

# make region
region <- make_region(sp.code, buffer = buffer, 
                      sub = region.sub, overwrite = T, continuous = F,
                      rm.clumps = T, lon.lo = lon.lo, lon.hi = lon.hi, lat.lo = lat.lo, lat.hi = lat.hi)
# cells in region$sp.grid will be removed if covariates have NA there
```


# Set up cross validation blocks ----
sb <- cv_spatial(x = region$sp.grid, 
                 rows_cols = c(block.rows, block.cols),
                 k = block.folds, 
                 hexagon = F, 
                 selection = "systematic", 
                 biomod2 = F, 
                 plot = F, 
                 report = F)


# Set up training and test sets based on cross validation blocks
if (block.out == "none") {
  test.i <- c()
  train.i <- region$sp.grid$conus.grid.id
  
} else {
  block1 <- sb$blocks %>%
    filter(folds == block.out)
  
  # find grid.ids for test block, everything else is train
  test.i1 <- st_intersection(region$sp.grid, block1)
  test.i <- st_intersection(region$sp.grid, block1) %>%
    pull(conus.grid.id) %>%
    unique()
  train.i <- filter(region$sp.grid, conus.grid.id %in% test.i == F) %>%
    pull(conus.grid.id)
  
}



# Covariate data ----
covar <- load_covar(sp.code,
                    region)

# read in covs.z
covs.z <- c(covs.lin, covs.quad, covs.int.factor)
if ("" %in% covs.z) {
  covs.z <- covs.z[-which(covs.z == "")]  
}
# 1/16/25: ROM added
if (NA %in% covs.z) {
  covs.z <- covs.z[-which(is.na(covs.z))]
}



# Add additional/derived covariates -----
# If you add a new covariate, add a row to data/covariate-labels.csv with a label
if (sp.code == "ANMI") { # add in additional ANMI covariates
 
  inv <- read_rds("data/species/ANMI/invsp-grid.rds") %>%
          mutate(CRAY01 = case_when(CRAY_huc10 > 0 ~ 1,
                                    CRAY_huc10 == 0 ~ 0),
                 BULL01 = BULL_huc10,
                 FISH01 = case_when(FISH_huc10 > 0 ~ 1,
                                    FISH_huc10 == 0 ~ 0),
                 nonnative = CRAY01 + BULL01 + FISH01)
  flow <- read_rds("data/species/ANMI/flow-grid.rds")
  clade <- read_rds("data/species/ANMI/clade-grid.rds")
  cwd <- read_rds("data/species/ANMI/cwd-grid.rds")
  rip <- read_rds("data/species/ANMI/riparian-grid.rds")
  rsvr <- read_rds("data/species/ANMI/rsvr-grid.rds")
  aet <- read_rds("data/species/ANMI/aet-grid.rds")
  perm <- read_rds("data/species/ANMI/waterperm-grid.rds")
  
  covar <- left_join(covar, select(inv, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(flow, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(perm, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(aet, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(cwd, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(rip, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(rsvr, !sp.grid.id), by = c("conus.grid.id")) %>%
            mutate(riparian = stream_forest) %>%
            mutate(sqrt.mean.spring.discharge = sqrt(mean.spring.discharge),
                   curt.mean.spring.discharge = mean.spring.discharge^(1/3)) %>%
            left_join(select(ungroup(clade), !sp.grid.id), by = c("conus.grid.id"))
    
  rm <- which(complete.cases(covar[,covs.z]) == F)
  if (length(rm) > 0) {
    covar <- covar[-rm,]
    region$sp.grid <- region$sp.grid[-rm,]
  }
  
} else if (sp.code == "RACA") {
  
  rain <- read_rds("data/species/RACA/rain-grid.rds")
  
  covar <- left_join(covar, select(rain, !sp.grid.id), by = c("conus.grid.id")) %>%
            mutate(sqrtarea_small = sqrt(area_small),
                   sqrtarea_medium = sqrt(area_medium))
  
  rm <- which(complete.cases(covar[,covs.z]) == F)
  if (length(rm) > 0) {
    covar <- covar[-rm,]
    region$sp.grid <- region$sp.grid[-rm,]
  }
  
  
} else if (sp.code == "GPOR") {
  
  cwd <- read_rds("data/species/GPOR/cwd-grid.rds")
  rain <- read_rds("data/species/GPOR/rain-grid.rds")
  soilwater <- read_rds("data/species/GPOR/soilwater-grid.rds")
  climate_velocity <- read_rds("data/species/GPOR/climate_velocity.rds")
  
  covar <- left_join(covar, select(cwd, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(rain, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(soilwater, !sp.grid.id), by = c("conus.grid.id")) %>%
            left_join(select(climate_velocity, !sp.grid.id), by = c("conus.grid.id"))
  
  rm <- which(complete.cases(covar[,covs.z]) == F)
  if (length(rm) > 0) {
    covar <- covar[-rm,]
    region$sp.grid <- region$sp.grid[-rm,]
  }
  
} else if (sp.code == "PLSE") {
  
  if (buffer == 1) regs <- c("east", "west", "west", "west", "west")
  if (buffer == 100000) regs <- c("east", "west", "west")
  reg <- region$region %>%
    st_cast("POLYGON") %>%
    mutate(region = regs)
  
  grid <- region$sp.grid %>%
    st_intersection(reg) %>%
    st_drop_geometry()
  
  covar <- covar %>%
    mutate(N_all = N + NW + NE) %>%
    full_join(grid, by = c("conus.grid.id", "sp.grid.id"))
  
  rm <- which(complete.cases(covar[,covs.z]) == F)
  if (length(rm) > 0) {
    covar <- covar[-rm,]
    region$sp.grid <- region$sp.grid[-rm,]
  }
  
}


# Scale covariates ----
covar_unscaled <- covar

# scale numeric cols
numcols <- sapply(covar, is.numeric)
numcols <- which(numcols)
numcols <- numcols[-which(numcols == 2)] # exclude sp.grid, which is numeric but shouldn't be scaled
covar[,numcols] <- sapply(covar[,numcols], scale_this)

# remove covariates that are correlated > 0.4
if (check.covs == T){
  covs.rm <- select_covar(covs.z, threshold = 0.4)
  covs.lin <- covs.lin[-which(covs.lin %in% covs.rm)]
  covs.quad <- covs.quad[-which(covs.quad %in% covs.rm)]
  
  covs.z <- c(covs.lin, covs.quad)
}

if (length(covs.z) < 3) {
  stop("There are fewer than 3 covariates remaining! This probably isn't a good model")
}



# Save covariate plots
if (block.out == "none") {
  
  # Process covs
  plot_covar(covar,
             covs.z,
             covs.int.factor,
             region,
             path = out.dir,
             type = "a_process")
  
  cor_covar(covar, covs.z, covs.int.factor, path = out.dir,
            type = "a_process", threshold = 0.25)
  
  # iNat covs
  plot_covar(covar,
             covs.inat,
             covs.int.factor = NA,
             region,
             path = out.dir,
             type = "b_iNat")
  
  if (length(covs.inat) > 1) cor_covar(covar, covs.inat, covs.int.factor = NA, path = out.dir, type = "b_iNat")
  
  # PO covs
  plot_covar(covar,
             covs.PO,
             covs.int.factor = NA,
             region,
             path = out.dir,
             type = "c_PO")
  
  if (length(covs.PO) > 1) cor_covar(covar, covs.PO, covs.int.factor = NA, path = out.dir, type = "c_PO")
}



# Add quadratic covariates and interactions ----
if (length(covs.quad) > 0 & paste0(covs.quad, collapse = "") != "") {
  for (c in 1:length(covs.quad)) {
    covar[,paste0(covs.quad[c], "2")] <- covar[,covs.quad[c]] * covar[,covs.quad[c]]
    covs.z <- c(covs.z, paste0(covs.quad[c], "2"))
  }
}

# and interactions
if (length(covs.int.factor) == 1 & is.na(covs.int.factor) == F) {
  
  # pivot factor wider to get dummy cols
  covar[,covs.int.factor] <- paste0(covs.int.factor, ".", covar[,covs.int.factor])
  covar <- covar %>%
    mutate(fact = 1) %>%
    pivot_wider(names_from = all_of(covs.int.factor), values_from = fact, values_fill = 0)
  
  covs.z <- c(covs.z, colnames(covar)[grep(paste0(covs.int.factor, "."), colnames(covar))])
  
  for (j in 1:length(covs.int.cont)) {
    # add interaction
    tmp <- add_int_cols(covar, int.factor = covs.int.factor, int.cont = covs.int.cont[j])
    covar <- tmp$covar
    covs.z <- c(covs.z, tmp$names)
  }
  
  # remove reference level
  rm <- grep(reference, covs.z)
  covs.z <- covs.z[-rm]
  
  # remove original column name
  rm <- grep(covs.int.factor, covs.z)[1]
  covs.z <- covs.z[-rm]
}



# Make gridkey ----
gridkey <- select(region$sp.grid, conus.grid.id) %>%
  st_drop_geometry() %>%
  mutate(grid.id = 1:nrow(.),
         group = case_when(conus.grid.id %in% train.i ~ "train",
                           conus.grid.id %in% test.i ~ "test"))

# Make spatkey ----
if (coarse.grid == T) {
  spatRegion <- suppressWarnings(make_spatkey(region$sp.grid))
  
  # Print spatial grid
  if (block.out == 'none') {
    pl <- ggplot(spatRegion$spat.grid) + geom_sf() + theme_bw()
    ggsave(pl, file = paste0(out.dir, "2_inputmap-e_spatGrid.jpg"), height = 8, width = 10)
  }
} else {
  spatRegion <- NULL
}





# Species data ----
species.data <- load_species_data(sp.code, 
                                  region, 
                                  filter.region = T,
                                  spat.bal = T,
                                  coordunc = 1000,
                                  coordunc_na.rm = T,
                                  data.prop = 1,
                                  year.start = year.start,
                                  year.end = year.end,
                                  keep.conus.grid.id = train.i) # only load data that are in correct grid cells

# ### Plot species data ----
# if (block == "none") {
#   tit <- ", full model"
# } else {
#   tit <- paste0(", excluding block ", block)
# }
# 
# pl <- map_species_data(sp.code = sp.code,
#                        region = region,
#                        species.data = species.data,
#                        year.start = year.start,
#                        year.end = year.end,
#                        plot = "samples",
#                        plot.region = T,
#                        sp.grid = sp.grid,
#                        details = T,
#                        title = paste0(common, " (", sp.code, ")", tit))
# ggsave(pl, file = paste0(out.dir, "2_inputmap-b_data-", blockname, "-details.jpg"), height = 8, width = 10)
# 
# pl <- map_species_data(sp.code = sp.code,
#                        region = region,
#                        species.data = species.data,
#                        year.start = year.start,
#                        year.end = year.end,
#                        plot = "samples",
#                        plot.region = T,
#                        sp.grid = sp.grid,
#                        details = F,
#                        title = paste0(common, " (", sp.code, ")", tit))
# ggsave(pl, file = paste0(out.dir, "2_inputmap-a_data-", blockname, ".jpg"), height = 8, width = 10)
# 
# 
# if (block != "none") {
#   
#   pl <- map_species_data(sp.code = sp.code,
#                          region = region,
#                          species.data = species.data,
#                          year.start = year.start,
#                          year.end = year.end,
#                          plot = "samples",
#                          plot.blocks = T,
#                          blocks = sb$blocks[which(sb$blocks$folds == block),],
#                          plot.region = T,
#                          sp.grid = sp.grid,
#                          details = T,
#                          title = paste0(common, " (", sp.code, ")", tit))
#   ggsave(pl, file = paste0(out.dir, "2_inputmap-d_blocks-", blockname, "-details.jpg"), height = 8, width = 10)
#   
#   pl <- map_species_data(sp.code = sp.code,
#                          region = region,
#                          species.data = species.data,
#                          year.start = year.start,
#                          year.end = year.end,
#                          plot = "samples",
#                          plot.blocks = T,
#                          blocks = sb$blocks[which(sb$blocks$folds == block),],
#                          plot.region = T,
#                          sp.grid = sp.grid,
#                          details = F,
#                          title = paste0(common, " (", sp.code, ")", tit))
#   ggsave(pl, file = paste0(out.dir, "2_inputmap-c_blocks-", blockname, ".jpg"), height = 8, width = 10)
#   
#   
# } else {
#   pl <- map_species_data(sp.code = sp.code,
#                          region = region,
#                          species.data = species.data,
#                          year.start = year.start,
#                          year.end = year.end,
#                          plot = "samples",
#                          plot.blocks = T,
#                          blocks = sb$blocks,
#                          plot.region = T,
#                          sp.grid = sp.grid,
#                          details = T,
#                          title = paste0(common, " (", sp.code, ")", tit))
#   ggsave(pl, file = paste0(out.dir, "2_inputmap-d_blocks-", blockname, "-details.jpg"), height = 8, width = 10)
# 
#   pl <- map_species_data(sp.code = sp.code,
#                          region = region,
#                          species.data = species.data,
#                          year.start = year.start,
#                          year.end = year.end,
#                          plot = "samples",
#                          plot.blocks = T,
#                          blocks = sb$blocks,
#                          plot.region = T,
#                          sp.grid = sp.grid,
#                          details = F,
#                          title = paste0(common, " (", sp.code, ")", tit))
#   ggsave(pl, file = paste0(out.dir, "2_inputmap-c_blocks-", blockname, ".jpg"), height = 8, width = 10)
#   
# }




# NIMBLE ----

# if (sp.code == "ANMI") {
#   
#   # In case datasets use the same site.ids, reassign them here
#   calls <- bind_rows(species.data$obs$`NMGFD Call surveys`, species.data$obs$`AZGFD Call surveys`) %>%
#     group_by(source, site.id) %>%
#     mutate(site.id = cur_group_id(),
#            source = "AZGFD/NMGFD Call Surveys") %>%
#     ungroup()
#   songmeter <- bind_rows(species.data$obs$`NMGFD Song`, species.data$obs$`AZGFD Songmeter`) %>%
#     group_by(source, site.id) %>%
#     mutate(site.id = cur_group_id(),
#            source = "AZGFD/NMGFD Songmeter") %>%
#     ungroup()
#   edna <- bind_rows(species.data$obs$`USGS eDNA`, species.data$obs$`USGS SNV eDNA`) %>%
#     group_by(source, site.id) %>%
#     mutate(site.id = cur_group_id(),
#            source = "USGS eDNA") %>%
#     ungroup()
#   ves <- bind_rows(species.data$obs$`USGS VES`, species.data$obs$`USGS SNV VES`) %>%
#     group_by(source, site.id) %>%
#     mutate(site.id = cur_group_id(),
#            source = "USGS VES") %>%
#     ungroup()
#   
#   
#   species.data.old <- species.data
#   
#   species.data$obs$`NMGFD Call surveys` <- NULL
#   species.data$obs$`AZGFD Call surveys` <- calls
#   
#   species.data$obs$`NMGFD Song` <- NULL
#   species.data$obs$`AZGFD Songmeter` <- songmeter
#   
#   species.data$obs$`USGS SNV eDNA` <- NULL
#   species.data$obs$`USGS eDNA` <- edna
#   
#   species.data$obs$`USGS SNV VES` <- NULL
#   species.data$obs$`USGS VES` <- ves
# }




sp.data <- sppdata_for_nimble(sp.code,
                              species.data,
                              covar,
                              covs.inat = covs.inat,
                              covs.PO = covs.PO,
                              CMR.to.count = T,
                              DND.maybe = 1,
                              keep.conus.grid.id = train.i) # get rid of PO cells that are in the wrong grid cells


### Data/constants ----
tmp <- data_for_nimble(sp.data, covar = covar, covs.z,
                       sp.auto = sp.auto, coarse.grid = coarse.grid, sp.grid = region$sp.grid,
                       area = T, process.intercept = process.intercept,
                       gridkey = gridkey, spatRegion= spatRegion)

data <- tmp$data
constants <- tmp$constants


if (sp.code == "ANMI") {

  # Need to replace NMGFD datasets with Bernoulli instead of occupancy
  for (s in 1:length(species.data$obs)) {
    nm <- constants[[paste0("name", s)]]
    nm <- gsub(" .*", "", nm)

    if (nm %in% c("NMGFD", "AZGFD")) {
      #print(constants[[paste0("nVisitsV", s)]])
      constants[[paste0("nVisitsV", s)]] <- 1
    }
  }

}

# if (sp.code == "ANMI") {
#   
#   constants[["nVisitsV", 2]] <- 1
#   constants[["nVisitsV", 3]] <- 1
#   
# }


if (model == "proj08nostate") {
  if (sp.code == "EBIS") {
    data$Xw3$WV <- NULL
    data$Xw3$VT <- NULL
    data$S3 <- NULL
    
    constants$nCovW3 <- ncol(data$Xw3)
  } else {
    stop("Don't know which states to remove")
  }
  
  rm.state <- T
} else {
  rm.state <- F
}



### Code ----
code <- nimble_code(data,
                    constants, 
                    path = out.dir,
                    sp.auto = sp.auto, 
                    coarse.grid = coarse.grid,
                    Bpriordist = Bpriordist, Bpriorvar1 = Bpriorvar1, Bpriorvar2 = Bpriorvar2,
                    block.out = block.out,
                    min.visits.incl = 3, 
                    zero_mean = zero_mean,
                    rm.state = rm.state)

### Initial values ----
inits <- function(x){nimble_inits(data,
                                  constants,
                                  sp.auto = sp.auto,
                                  seed = x)}

### Parameters ----
params <- nimble_params(data,
                        constants,
                        lambda = T,
                        # psi = T,
                        XB = T,
                        sp.auto = sp.auto)
                        #project = project)



end1 <- Sys.time() - start1

# Remove local and block in case the setup is run locally but the model is fit on the HPC.
rm(list=c('local','block','args'))

# Remove all functions (will reload in a later step)
fxns <- as.character(lsf.str())
fxns <- fxns[-which(fxns == 'inits')] # Keep inits fxn in environment
rm(list=fxns)

# Save environment and full set up
save.image(paste0(out.dir, "setup_",block.out,".Rdata"))


# End script - proceed to 2-MVPv1-test.R
